Compositional Quantitative Resource Analysis of C Programs

### 2014-06-17 paper/main.tex
change notations
more work on section 3

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid E \gets E
\mid E \gets f(E^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

The expressions are left abstract in this presentation.  To
present both the semantics and the program logic, we simply
need an evaluation function that maps any $e \in E$ to
a value $\llbracket e \rrbracket_H$ using $H$ as current
heap.

@@

Values computed by evaluating expressions can be integers, or
memory locations. In our semantics, integers have infinite
precision.  While the program logic would perform as well on
machine integers, our automatic analysis would be unsound.
%
\begin{align*}
V &:= \code{int}~i \mid \code{ptr}~b
\end{align*}

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  To account for all kinds of resources
we leave it as a parameter of the semantics to be specified
by the user.  This parameter, called a metric, is a tuple of rational
numbers with the following form
$$
(
M_a, M_s, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q
)
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

The general form of an evaluation judgement is
$(H, S) \Downarrow_{(q,q')} X$.  It is similar to the judgement
presented in the introducing section appart from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid by
turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
This injection is used implicitely in the semantics.

@@

\Rule{S:Abort}
{ }
{ (H, S) \Downarrow_{(0,0)} \circ }

\and \Rule{S:Assert}
{ {\rm istrue}~\llbracket e \rrbracket_H}
{ (H, \code{assert}~e) \Downarrow_{M_a} \code{seq}~H }

\and \Rule{S:Skip}
{ }
{ (H, \code{skip}) \Downarrow_{M_s} \code{seq}~H }

\and \Rule{S:Break}
{ }
{ (H, \code{break}) \Downarrow_{M_b} \code{brk}~H }

\and \Rule{S:Return}
{ }
{ (H, \code{return}~e) \Downarrow_{M_r} \code{ret}~H~\llbracket e \rrbracket_H }

\and \Rule{S:Update}
{ \llbracket e \rrbracket_H = \code{ptr}~b }
{ (H, e \gets e') \Downarrow_{M_u} \code{seq}~H[b \mapsto \llbracket e' \rrbracket_H] }

\and \Rule{S:FunCall}
{ \Delta~f = ((a_1, \dots, a_n), S)
\\ \llbracket e \rrbracket_H = \code{ptr}~b
\\ (H +\!\!\!+ (a_i \mapsto \llbracket e_i \rrbracket_H)_i, S) \Downarrow_{(q,q')} \code{ret}~H'~v
}
{ (H, e \gets f(e_1, \dots, e_n))
\Downarrow^r_{M_f\cdot(q,q')} \code{seq}~(H'\,\backslash\,(a_i)_i)[b \mapsto v] }

\and \Rule{S:LoopSeq}
{ (H, S) \Downarrow_{(q_1, q_1')} \code{seq}~H'
\\ (H', \code{loop}~S) \Downarrow_{(q_2, q_2')} X }
{ (H, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} X }

\and \Rule{S:LoopBrk}
{ (H, S) \Downarrow_{(q, q')} \code{brk}~H' }
{ (H, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~H' }

\and \Rule{S:LoopAbort}
{ (H, S) \Downarrow_{(q, q')} X
\\ \forall H.~X \neq \code{seq}~H \land X \neq \code{brk}~H }
{ (H, \code{loop}~S) \Downarrow_{(q, q')} X }

\and \Rule{S:IfThen}
{  (H, S_1) \Downarrow_{(q, q')} X
\\ {\sf istrue}~\llbracket C \rrbracket_H
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^1 \cdot (q, q')} X }

\and \Rule{S:IfElse}
{  (H, S_2) \Downarrow_{(q, q')} X
\\ {\sf isfalse}~\llbracket C \rrbracket_H
}
{ (H, \code{if}(C)~S_1~\code{else}~S_2) \Downarrow_{M_c^2 \cdot (q, q')} X }

\and \Rule{S:Seq}
{  (H, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~H'
\\ (H', S_2) \Downarrow_{(q_2, q'_2)} X
}
{ (H, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} X }

\and \Rule{S:SeqAbort}
{ (H, S_1) \Downarrow_{(q,q')} X
\\ \forall H.~X \neq \code{seq}~H}
{ (H, S_1; S_2) \Downarrow_{(q,q')} X }

@@

Heaps manipulated by the operational semantics judgements are
simple maps from variable names to integers.  We designate by
$H[x \mapsto v]$ the heap $H$ updated to map the variable name
$x$ to the number $v$.
%
The evaluation of expressions is trivial using heaps.
\begin{mathpar}
\llbracket x \rrbracket_H = H(x)
\and
\llbracket x \pm y \rrbracket_H = H(x) \pm H(y)
\end{mathpar}

In the evaluation judgement $(H, S) \Downarrow_{(q, q')} X$, the result $X$ can
be either a new heap, which means the execution terminated successfully
or it can be the \emph{busy} result $\circ$.  This indicates an early
termination, it can be triggered at any point using the rule {\sc S:Abort}
and bubbles up through the execution context.
Sometimes there is no other choice than to use this latter rule, for
example when an assertion fails.  In that case the side condition of
{\sc S:Assert} is not met and the only rule that can possibly be applied
is {\sc S:Abort}.

>>>>

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid x \gets E
\mid x \gets f(E^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

The expressions are left abstract in this presentation.  To
present both the semantics and the program logic, we simply
need an evaluation function that maps any $e \in E$ to
a value $\llbracket e \rrbracket_\state$ using $\state$ as current
state.

@@

The operational semantics uses a finite map $\state$ from
variable names to integers.  This map is represented as
an associative list to model the semantics of the stack of
our language. It can be extended with new bindings
(hiding older bindings with the same name), shrinked by
forgetting a list of bindings, and updated by changing one
binding.  We note $(\state, x_1 \mapsto v_1, \dots, x_n \mapsto v_n)$
an extended heap, $\state~\backslash~(x_1, \dots, x_n)$ a
state where bindings of $x_1, \dots, x_n$ were removed and
$\state[x \mapsto v]$ an updated state.

To formally describe the cost of operation of a given
program, our semantics tracks resource consumption while
executing.  The semantics is oblivious to the resource of
interest for the user of our system.  We acheive this
independence by parameterizing the semantics by a resource
\emph{metric}, it is a tuple of rational numbers with the
following form
$$
(
M_a, M_s, M_b, M_r, M_u, M_f,
M_l,
M_c^1, M_c^2,
M_q
).
$$
Each of these rational numbers indicates the amount of resource
consumed by an operation.  If one of these numbers is negative,
it means that resources are released by the program.

The general form of an evaluation judgement is
$(\state, S) \Downarrow_{(q,q')} O$.  It is similar to the judgement
presented in the introducing section appart from the fact that
two resource numbers are tracked instead of one.  Intuitively
$(q, q')$ means that $q$ resources units are necessary to execute
the program and that $q'$ units are available after its execution.
For example, $\code{alloc}(3)$ consumes 3 heap cells, so
the pair resulting from its evaluation would be $(3, 0)$; for
$\code{free}(3)$, the pair would be $(0, 3)$.  When these
statements are sequenced with \code{alloc} first, then \code{free},
the evaluation yields $(3, 3)$.  These resource pairs can
be composed with the following operation
$$
(q_1, q'_1) \cdot (q_2, q'_2) =
\left\{
\begin{array}{ll}
(q_1, q'_2 + q'_1 - q_2) & \mbox{if } q'_1 \ge q_2 \\
(q_1 + q_2 - q'_1, q'_2) & \mbox{if } q'_1 < q_2
\end{array}
\right.
.
$$
This operation can be seen as the sequential composition of
two resource consumptions, note that it is not a commutative
operation, the same way resource freeing and allocation do not
commute.
With $(0, 0)$ as a neutral element, the above composition defines
a monoid structure on $\mathbb Q_0^+ \times \mathbb Q_0^+$.
It is possible to inject $q \in \mathbb Q$ in this monoid by
turning it into $(q, 0)$ if $q \ge 0$ and $(0, -q)$ otherwise.
This injection is used implicitely in the semantics.

@@

\Rule{S:Abort}
{ }
{ (\state, S) \Downarrow_{(0,0)} \circ }

\and \Rule{S:Assert}
{ {\rm istrue}~\llbracket e \rrbracket_\state}
{ (\state, \code{assert}~e) \Downarrow_{M_a} \code{seq}~\state }

\and \Rule{S:Skip}
{ }
{ (\state, \code{skip}) \Downarrow_{M_s} \code{seq}~\state }

\and \Rule{S:Break}
{ }
{ (\state, \code{break}) \Downarrow_{M_b} \code{brk}~\state }

\and \Rule{S:Return}
{ }
{ (\state, \code{return}~e) \Downarrow_{M_r}
  \code{ret}~\state~\llbracket e \rrbracket_\state
}

\and \Rule{S:Update}
{ }
{ (\state, x \gets e) \Downarrow_{M_u}
  \code{seq}~\state[x \mapsto \llbracket e \rrbracket_\state]
}

\and \Rule{S:FunCall}
{ \Delta~f = (x_1, \dots, x_n, S) \\
  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
}

\and \Rule{S:LoopSeq}
{ (\state, S) \Downarrow_{(q_1, q_1')} \code{seq}~\state'
\\ (\state', \code{loop}~S) \Downarrow_{(q_2, q_2')} O }
{ (\state, \code{loop}~S) \Downarrow_{(q_1, q_1') \cdot M_l \cdot (q_2, q_2')} O }

\and \Rule{S:LoopBrk}
{ (\state, S) \Downarrow_{(q, q')} \code{brk}~\state' }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} \code{seq}~\state' }

\and \Rule{S:LoopAbort}
{ (\state, S) \Downarrow_{(q, q')} O
\\ \forall \state.~O \neq \code{seq}~\state \land O \neq \code{brk}~\state }
{ (\state, \code{loop}~S) \Downarrow_{(q, q')} O }

\and \Rule{S:IfThen}
{  (\state, S_1) \Downarrow_{(q, q')} O
\\ {\sf istrue}~\llbracket C \rrbracket_\state
}
{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^1 \cdot (q, q')} O
}

\and \Rule{S:IfElse}
{  (\state, S_2) \Downarrow_{(q, q')} O
\\ {\sf isfalse}~\llbracket C \rrbracket_\state
}
{ (\state, \code{if}(C)~S_1~\code{else}~S_2)
  \Downarrow_{M_c^2 \cdot (q, q')} O
}

\and \Rule{S:Seq}
{  (\state, S_1) \Downarrow_{(q_1, q'_1)} \code{seq}~\state'
\\ (\state', S_2) \Downarrow_{(q_2, q'_2)} O
}
{ (\state, S_1; S_2) \Downarrow_{(q_1, q'_1) \cdot M_q \cdot (q_2, q'_2)} O }

\and \Rule{S:SeqAbort}
{ (\state, S_1) \Downarrow_{(q,q')} O
\\ \forall \state.~O \neq \code{seq}~\state}
{ (\state, S_1; S_2) \Downarrow_{(q,q')} O }

@@

In the evaluation judgement $(\state, S) \Downarrow_{(q, q')} O$,
the outcome $O$ specifies how the statement $S$ terminated.
$$
O := \circ
\mid \code{seq}~\sigma
\mid \code{brk}~\sigma
\mid \code{ret}~\sigma~v
$$
The \code{seq} outcome is the simple fallthrough outcome, for example
a \code{skip} statement will terminate with this outcome without changing
the heap as pictured in the rule {\sc S:Skip}.  The \code{brk} and
\code{ret} outcomes describe a statement that terminated with a break or
return statement.  In the latter case, a return value is also part of the
outcome.  The use of these different outcomes can be explained by considering
the loop statement rules:  If the inner statement breaks
then the loop exits using the rule {\sc S:LoopBrk}, if the inner statement
falls through another iteration is run using the rule {\sc S:LoopSeq},
and all other cases are handled in the rule {\sc S:LoopAbort}.

The special outcome $\circ$ indicates an early
termination, it can be triggered at any point using the rule {\sc S:Abort}
and bubbles up through the execution context.
Sometimes there is no other choice than to use this latter rule, for
example when an assertion fails.  In that case the side condition of
{\sc S:Assert} is not met and the only rule that can possibly be applied
is {\sc S:Abort}.
